---
title: "How to Access an API with an Example"
author: "Autumn Biggie"
date: "10/3/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this document, we'll walk through how to connect to an API, using an example. The API we'll be connecting to is the [OpenWeather API](https://openweathermap.org/api). Specifically, we'll be looking at current and forecast weather data from the [One Call API](https://openweathermap.org/api/one-call-api#history), one of the many APIs that openweathermap.org offers.  

## Preliminary Steps  

1.  Many APIs require you to access the data using a unique API key. To acquire your free API key, register [here](https://home.openweathermap.org/users/sign_up).  

2. The following packages will be necessary in order to connect with the API:  

    + `httr`  
    + `jsonlite`  

3. The following packages will be necessary in order to do some analyses after we load the data:

    + `tidyverse`  
    + `anytime`  

```{r API packages, echo = FALSE, message=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)
library(anytime)
```


## Function to Access API  

Here, I've written a function `weather.api` to be able to easily access the target API. The arguments are as follows:  

* `latitude`(required): latitude of the geographic location desired  
* `longitude`(required): longitude of the geographic location desired  
* `api.id`(required): input your unique API key here  
* `exclude`(optional): list any parts of the weather data you want to exclude from the results. This should be a comma-delimited list, with or without spaces. Case is unimportant. The options are:
    + `current`  
    + `minutely`  
    + `hourly`  
    + `daily`  
* `units` (optional): The unit of measurement in which values are returned. Options are:
    + `standard`  
    + `imperial`  
    + `metric`  


Note: All parameters should be in character format i.e. `latitude = "-46.05"`.  



```{r}
weather.api <- function(latitude, longitude, api.id, exclude = NULL, units = "metric") {
  
  #assign the individual pieces of the required url to their respective objects  
  base <- "https://api.openweathermap.org/data/2.5/onecall"
  lat1 <- "lat="
  lat2 <- latitude
  lon1 <- "lon="
  lon2 <- longitude
  exc1 <- "exclude="
  exc2 <- tolower(sub(" ", "", exclude)) #remove spaces and convert to lowercase  
  apid1 <- "appid="
  apid2 <- api.id
  units1 <- "units="
  units2 <- units
  
  #paste pieces together  
  lat <- paste(lat1, lat2, sep = "")
  lon <- paste(lon1, lon2, sep = "")  
  ifelse(!is.null(exc2), exc <- paste(exc1, exc2, sep = ""), exc <- "nullexclude")
  apid <- paste(apid1, apid2, sep = "")
  units <- paste(units1, units2, sep = "")
  
  #paste base and latitude pieces together  
  first <- paste(base, lat, sep = "?")
  
  #create the entire url by pasting all pieces together using delimiter "&", conditioning on if any information is excluded  
  ifelse(!is.null(exc2), my.url <- paste(first, lon, exc, apid, units, sep = "&"), my.url <- paste(first, lon, apid, units, sep = "&"))
  
  #access the Weather API using the URL assembled above
  weather.info <- GET(my.url)
  
  #convert the content of the accessed data from raw to character, then present in readable data frame format
  final <- weather.info$content %>% rawToChar() %>% fromJSON()

return(final)
}
```


## Exploratory Data Analysis  

Now that we can easily access the API, let's do some exploratory data analysis.  

I want to explore the weather data of five locations that lie roughly along the same line of longitude.   

[insert picture here]  

Locations:  

* 44.19'N, 69.47'W (Augusta, Maine, US)  
* 21.28'N, 71.08'W (Cockburn Town, Turks & Caicos Islands, UK)  
* 7.54'N, 72.3'W (Cucata, Colombia)  
* 55.05'S, 67.05'W (Puerto Toro, Chile)  
* 51.38'S, 69.13'W (Rio Gallegos, Argentina)  

```{r Maine API}
Maine <- weather.api(latitude = "44.19", longitude = "-69.47", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Maine.day <- Maine$daily

Maine.day
```

### Data Cleaning  

Let's convert all dates/times that are in unix form to a date/time stamp.  

```{r}
Maine.day$dt <- anydate(Maine.day$dt)
Maine.day$sunrise <- anytime(Maine.day$sunrise)
Maine.day$sunset <- anytime(Maine.day$sunset)
Maine.day$moonrise <- anytime(Maine.day$moonrise)
Maine.day$moonset <- anytime(Maine.day$moonset)
```

Convert the `dt` variable into three variables, `Year`, `Month`, `Day`, and then change `sunrise`, `sunset`, `moonrise`, and `moonset` to only include timestamps.  


```{r}
Maine.day <- Maine.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

#change sunrise, sunset, moonrise, and moonset to only include timestamps

Maine.day <- Maine.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Maine.day <- Maine.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Maine.day <- Maine.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Maine.day <- Maine.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

```

Lastly, combine the nested `temp` dataframe with `Maine.day` to create a new dataframe called `Maine.day1`, removing unwanted variables.  


```{r}
Maine.day1 <- data.frame(Maine.day, Maine.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds:rain, min, max) %>% rename(mintemp = min, maxtemp = max)

Maine.day1
```

Awesome! Remember to clean the data returned from the other locations too.  

```{r Turks & Caicos, echo=FALSE}
Turks <- weather.api(latitude = "21.28", longitude = "-71.08", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Turks.day <- Turks$daily

Turks.day$dt <- anydate(Turks.day$dt)
Turks.day$sunrise <- anytime(Turks.day$sunrise)
Turks.day$sunset <- anytime(Turks.day$sunset)
Turks.day$moonrise <- anytime(Turks.day$moonrise)
Turks.day$moonset <- anytime(Turks.day$moonset)

Turks.day <- Turks.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Turks.day <- Turks.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day <- Turks.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day <- Turks.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day <- Turks.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day1 <- data.frame(Turks.day, Turks.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds:rain, min, max) %>% rename(mintemp = min, maxtemp = max)

```

```{r Colombia, echo=FALSE}
Colombia <- weather.api(latitude = "7.54", longitude = "-72.3", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Colombia.day <- Colombia$daily

Colombia.day$dt <- anydate(Colombia.day$dt)
Colombia.day$sunrise <- anytime(Colombia.day$sunrise)
Colombia.day$sunset <- anytime(Colombia.day$sunset)
Colombia.day$moonrise <- anytime(Colombia.day$moonrise)
Colombia.day$moonset <- anytime(Colombia.day$moonset)

Colombia.day <- Colombia.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Colombia.day <- Colombia.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day <- Colombia.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day <- Colombia.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day <- Colombia.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day1 <- data.frame(Colombia.day, Colombia.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds:rain, min, max) %>% rename(mintemp = min, maxtemp = max)

```


```{r Chile, echo=FALSE}
Chile <- weather.api(latitude = "-55.05", longitude = "-67.05", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Chile.day <- Chile$daily

Chile.day$dt <- anydate(Chile.day$dt)
Chile.day$sunrise <- anytime(Chile.day$sunrise)
Chile.day$sunset <- anytime(Chile.day$sunset)
Chile.day$moonrise <- anytime(Chile.day$moonrise)
Chile.day$moonset <- anytime(Chile.day$moonset)

Chile.day <- Chile.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Chile.day <- Chile.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day <- Chile.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day <- Chile.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day <- Chile.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day1 <- data.frame(Chile.day, Chile.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds:uvi, min, max) %>% rename(mintemp = min, maxtemp = max)

```

```{r Argentina, echo=FALSE}
Argentina <- weather.api(latitude = "-51.38", longitude = "-69.13", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Argentina.day <- Argentina$daily

Argentina.day$dt <- anydate(Argentina.day$dt)
Argentina.day$sunrise <- anytime(Argentina.day$sunrise)
Argentina.day$sunset <- anytime(Argentina.day$sunset)
Argentina.day$moonrise <- anytime(Argentina.day$moonrise)
Argentina.day$moonset <- anytime(Argentina.day$moonset)

Argentina.day <- Argentina.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Argentina.day <- Argentina.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day <- Argentina.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day <- Argentina.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day <- Argentina.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day1 <- data.frame(Argentina.day, Argentina.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds:uvi, min, max) %>% rename(mintemp = min, maxtemp = max)

```

```{r return all datasets}
Turks.day1
Colombia.day1
Chile.day1
Argentina.day1
```


