---
title: "How to Access an API with an Example"
author: "Autumn Biggie"
date: "10/3/2021"
output: 
  github_document:
    html_preview: FALSE
    includes:
      in_header: header.html
always_allow_html: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this document, we'll walk through how to connect to an API, using an example. The API we'll be connecting to is the [OpenWeather API](https://openweathermap.org/api). Specifically, we'll be looking at current and forecast weather data from the [One Call API](https://openweathermap.org/api/one-call-api#history), one of the many APIs that openweathermap.org offers.  

## Preliminary Steps  

1.  Many APIs require you to access the data using a unique API key. To acquire your free API key, register [here](https://home.openweathermap.org/users/sign_up).  

2. The following packages will be necessary in order to connect with the API:  

    + `httr`  
    + `jsonlite`  

3. The following packages will be necessary in order to do some analyses after we load the data:

    + `tidyverse`  
    + `anytime`  
    + `ggplot2`
    + `chron`
    + `leaflet`

```{r API packages, echo = FALSE, message=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)
library(anytime)
library(ggplot2)
library(chron)
library(leaflet)
library(knitr)
```


## Function to Access API  

Here, I've written a function `weather.api` to be able to easily access the target API. The arguments are as follows:  

* `latitude`(required): latitude of the geographic location desired  
* `longitude`(required): longitude of the geographic location desired  
* `api.id`(required): input your unique API key here  
* `exclude`(optional): list any parts of the weather data you want to exclude from the results. This should be a comma-delimited list, with or without spaces. Case is unimportant. The options are:
    + `current`  
    + `minutely`  
    + `hourly`  
    + `daily`  
* `units` (optional): The unit of measurement in which values are returned. Options are:
    + `standard`  
    + `imperial`  
    + `metric`  


Note: All parameters should be in character format i.e. `latitude = "-46.05"`.  



```{r API access function}
weather.api <- function(latitude, longitude, api.id, exclude = NULL, units = "metric") {
  
  #assign the individual pieces of the required url to their respective objects  
  base <- "https://api.openweathermap.org/data/2.5/onecall"
  lat1 <- "lat="
  lat2 <- latitude
  lon1 <- "lon="
  lon2 <- longitude
  exc1 <- "exclude="
  exc2 <- tolower(sub(" ", "", exclude)) #remove spaces and convert to lowercase  
  apid1 <- "appid="
  apid2 <- api.id
  units1 <- "units="
  units2 <- units
  
  #paste pieces together  
  lat <- paste(lat1, lat2, sep = "")
  lon <- paste(lon1, lon2, sep = "")  
  ifelse(!is.null(exc2), exc <- paste(exc1, exc2, sep = ""), exc <- "nullexclude")
  apid <- paste(apid1, apid2, sep = "")
  units <- paste(units1, units2, sep = "")
  
  #paste base and latitude pieces together  
  first <- paste(base, lat, sep = "?")
  
  #create the entire url by pasting all pieces together using delimiter "&", conditioning on if any information is excluded  
  ifelse(!is.null(exc2), my.url <- paste(first, lon, exc, apid, units, sep = "&"), my.url <- paste(first, lon, apid, units, sep = "&"))
  
  #access the Weather API using the URL assembled above
  weather.info <- GET(my.url)
  
  #convert the content of the accessed data from raw to character, then present in readable data frame format
  final <- weather.info$content %>% rawToChar() %>% fromJSON()

return(final)
}
```


## Exploratory Data Analysis  

Now that we can easily access the API, let's do some exploratory data analysis.  

I want to explore the weather data of five locations that lie roughly along the same line of longitude.   

```{r map, echo=FALSE}
leaflet() %>% setView(lng = -80, lat = 0, zoom = 1) %>% addTiles() %>% addMarkers(c(-69.47, -71.08, -72.3, -67.05, -69.13), c(44.19, 21.28, 7.54, -55.05, -51.38), popup = c("Augusta, Maine, US", "Cockburn Town, Turks & Caicos Islands", "Cucata, Colombia", "Puerto Toro, Chile", "Rio Gallegos, Argentina"))
```
 

Locations:  

* 44.19'N, 69.47'W (Augusta, Maine, US)  
* 21.28'N, 71.08'W (Cockburn Town, Turks & Caicos Islands, UK)  
* 7.54'N, 72.3'W (Cucata, Colombia)  
* 51.38'S, 69.13'W (Rio Gallegos, Argentina) 
* 55.05'S, 67.05'W (Puerto Toro, Chile)  

```{r Maine API display, eval=FALSE, echo=TRUE}
Maine <- weather.api(latitude = "44.19", longitude = "-69.47", api.id = "input_your_key", exclude = "current, minutely, hourly")

Maine.day <- Maine$daily

Maine.day
```

```{r Maine API functional, echo=FALSE}
Maine <- weather.api(latitude = "44.19", longitude = "-69.47", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Maine.day <- Maine$daily

Maine.day
```

### Data Cleaning  

Let's convert all dates/times that are in unix form to a date/time stamp.  

```{r}
Maine.day$dt <- anydate(Maine.day$dt)
Maine.day$sunrise <- anytime(Maine.day$sunrise)
Maine.day$sunset <- anytime(Maine.day$sunset)
Maine.day$moonrise <- anytime(Maine.day$moonrise)
Maine.day$moonset <- anytime(Maine.day$moonset)
```

Convert the `dt` variable into three variables, `Year`, `Month`, `Day`, and then change `sunrise`, `sunset`, `moonrise`, and `moonset` to only include timestamps.  


```{r}
Maine.day <- Maine.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

#change sunrise, sunset, moonrise, and moonset to only include timestamps

Maine.day <- Maine.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Maine.day <- Maine.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Maine.day <- Maine.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Maine.day <- Maine.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

```

Lastly, combine the nested `temp` dataframe with `Maine.day` to create a new dataframe called `Maine.day1`, removing unwanted variables.  


```{r}
Maine.day1 <- data.frame(Maine.day, Maine.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds, pop, min, max) %>% rename(mintemp = min, maxtemp = max)

Maine.day1
```

Awesome! Remember to clean the data returned from the other locations too.  

```{r Turks & Caicos, echo=FALSE}
Turks <- weather.api(latitude = "21.28", longitude = "-71.08", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Turks.day <- Turks$daily

Turks.day$dt <- anydate(Turks.day$dt)
Turks.day$sunrise <- anytime(Turks.day$sunrise)
Turks.day$sunset <- anytime(Turks.day$sunset)
Turks.day$moonrise <- anytime(Turks.day$moonrise)
Turks.day$moonset <- anytime(Turks.day$moonset)

Turks.day <- Turks.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Turks.day <- Turks.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day <- Turks.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day <- Turks.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day <- Turks.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Turks.day1 <- data.frame(Turks.day, Turks.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds, pop, min, max) %>% rename(mintemp = min, maxtemp = max)

```

```{r Colombia, echo=FALSE}
Colombia <- weather.api(latitude = "7.54", longitude = "-72.3", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Colombia.day <- Colombia$daily

Colombia.day$dt <- anydate(Colombia.day$dt)
Colombia.day$sunrise <- anytime(Colombia.day$sunrise)
Colombia.day$sunset <- anytime(Colombia.day$sunset)
Colombia.day$moonrise <- anytime(Colombia.day$moonrise)
Colombia.day$moonset <- anytime(Colombia.day$moonset)

Colombia.day <- Colombia.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Colombia.day <- Colombia.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day <- Colombia.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day <- Colombia.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day <- Colombia.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Colombia.day1 <- data.frame(Colombia.day, Colombia.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds, pop, min, max) %>% rename(mintemp = min, maxtemp = max)

```


```{r Chile, echo=FALSE}
Chile <- weather.api(latitude = "-55.05", longitude = "-67.05", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Chile.day <- Chile$daily

Chile.day$dt <- anydate(Chile.day$dt)
Chile.day$sunrise <- anytime(Chile.day$sunrise)
Chile.day$sunset <- anytime(Chile.day$sunset)
Chile.day$moonrise <- anytime(Chile.day$moonrise)
Chile.day$moonset <- anytime(Chile.day$moonset)

Chile.day <- Chile.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Chile.day <- Chile.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day <- Chile.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day <- Chile.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day <- Chile.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Chile.day1 <- data.frame(Chile.day, Chile.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds, pop, min, max) %>% rename(mintemp = min, maxtemp = max)

```

```{r Argentina, echo=FALSE}
Argentina <- weather.api(latitude = "-51.38", longitude = "-69.13", api.id = "27667529a1629f208f81fded8f7552af", exclude = "current, minutely, hourly")

Argentina.day <- Argentina$daily

Argentina.day$dt <- anydate(Argentina.day$dt)
Argentina.day$sunrise <- anytime(Argentina.day$sunrise)
Argentina.day$sunset <- anytime(Argentina.day$sunset)
Argentina.day$moonrise <- anytime(Argentina.day$moonrise)
Argentina.day$moonset <- anytime(Argentina.day$moonset)

Argentina.day <- Argentina.day %>% separate(dt, c("Year", "Month", "Day"), sep = "-", convert = TRUE, remove = TRUE)

Argentina.day <- Argentina.day %>% separate(sunrise, c("Date", "Sunrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day <- Argentina.day %>% separate(sunset, c("Date", "Sunset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day <- Argentina.day %>% separate(moonrise, c("Date", "Moonrise"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day <- Argentina.day %>% separate(moonset, c("Date", "Moonset"), sep = " ", remove = TRUE) %>% subset(select = -Date)

Argentina.day1 <- data.frame(Argentina.day, Argentina.day$temp) %>% select(Year:moon_phase, pressure:wind_gust, clouds, pop, min, max) %>% rename(mintemp = min, maxtemp = max)

```


Now, let's combine all datasets into one called `weather`, creating a new variable called `location`.  

```{r}
Maine.day1 <- Maine.day1 %>% mutate(location = "Maine, US")
Turks.day1 <- Turks.day1 %>% mutate(location = "Turks & Caicos")
Colombia.day1 <- Colombia.day1 %>% mutate(location = "Colombia")
Chile.day1 <- Chile.day1 %>% mutate(location = "Chile")
Argentina.day1 <- Argentina.day1 %>% mutate(location = "Argentina")

weather <- rbind(Maine.day1, Turks.day1, Colombia.day1, Argentina.day1, Chile.day1) %>% relocate(location, .before = Year)

```


### Create New Variables  

I'm interested in converting `humidity` and `clouds` into categorical variables with different levels.  

I'll begin with humidity. Lets say that if humidity is less than or equal to 60, there is low humidity, if 60 < humidity is less than or equal to 80, there is medium humidity, and if humidity > 80, there is high humidity.  

```{r humidity status}
weather <- weather %>% mutate(humidity.status = as.factor(ifelse(humidity > 80, "High", ifelse(humidity >60, "Medium", "Low"))))

weather$humidity.status <- ordered(weather$humidity.status, levels = c("Low", "Medium", "High"))

```


Great! Now let's look at `clouds`. If clouds is less than or equal to 25, then cloud coverage is low. If clouds > 75 then cloud coverage is high, and anything in between is medium.  


```{r cloud coverage}
weather <- weather %>% mutate(cloud.coverage = as.factor(ifelse(clouds > 75, "High", ifelse(clouds > 25, "Medium", "Low"))))

weather$cloud.coverage <- ordered(weather$cloud.coverage, levels = c("Low", "Medium", "High"))

```


Here's our cleaned dataset:  
```{r final dataset}
weather
```


### Contingency Tables  

With the new categorical variables I've created, let's create some contingency tables. `tabz1` will show the counts of observations within each level combination of `cloud.coverage` and `humidity.status`.  


```{r contingency table 1}
tabz1 <- table(weather$humidity.status, weather$cloud.coverage, deparse.level = 2)

tabz1
```

It looks like there were 10 forecast observations that predicted high cloud coverage and low humidity, and 12 observations that predicted high cloud coverage and medium humidity. There weren't many observations that predicted high humidity, but it appears that when high humidity was predicted, it was most often coupled with high cloud coverage.  


`tabz2` will show the counts of days within each level combination of `cloud.coverage` and `humidity.status`, separated by `location`.  


```{r contingency table 2}
tabz2 <- table(weather$humidity.status, weather$cloud.coverage, weather$location, deparse.level = 2)

tabz2
```

This table gives us an idea of how the relationship between humidity status and cloud coverage can change based on location. An observation to note is that Maine and Argentina, locations close to the poles, most often have low humidity in this dataset. However, the low humidity is more frequently coupled with high cloud coverage in Argentina and low cloud coverage in Maine. I'd be interested to find out why this is the case, or if the data only appears this way because we're not working with many observations. Also, it's interesting that Chile, the location closest to the south pole, has a range of humidity although it's close to the Argentina location.  


### Numerical Summaries  

Now that we've explored our categorical variables, let's take a look at some numeric summaries. 

The table below lists the average maximum temperature, average minimum temperature, their respective standard deviations, and the interquartile range for each location.  


```{r temp numerical summary}
weather$location <- ordered(weather$location, levels = c("Chile", "Argentina", "Colombia", "Turks & Caicos", "Maine, US")) #convert location into an ordered variable

weather %>% group_by(location) %>% summarise(avghigh = mean(maxtemp), avglow = mean(mintemp), sdhigh = sd(maxtemp), sdlow = sd(mintemp), IQR = IQR(maxtemp))
```

It's clear that the average high and average low peak near the middle of the globe, with greater variation near the poles. These numbers could not only be affected by latitude, but also by elevation.  


Now we'll look at the average humidity forecasted for each location, as well as their standard deviations.  


```{r}
weather %>% group_by(location) %>% summarise(avg_humidity = mean(humidity), sd.humidity = sd(humidity))
```

Here we see a clear relationship between location (organized by latitude) and humidity. The locations nearest to the poles generally have the lowest humidity, but the Turks and Caicos Islands as well as the Chile location have the highest humidity, most likely because they are close to or in the ocean.  


### Data Visualization  

Now I'll use some tools for visualizing the data we've collected from the API.  

Let's visualize the second contingency table we made above using a bar graph with `humidity.status` and `cloud.coverage`, separating the results by `location`.  

```{r bar graph, message=FALSE, warning=FALSE}
sum.tab <- weather %>% group_by(location, humidity.status, cloud.coverage) %>% summarise(count = n())

g6 <- ggplot(sum.tab, aes(x = humidity.status, y = count))

g6 + geom_bar(aes(fill = cloud.coverage), stat = "identity", position = "dodge") + facet_wrap(~ location) + labs(title = "8 Day Forecast: Humidity Status vs. Cloud Coverage by Location", x = "Humidity Status") + scale_fill_discrete(name = "Cloud Coverage") + theme(axis.text.x = element_text(angle = 45, vjust = .8, hjust = 1))
```

Here we can see the observation we made earlier about Maine and Argentina having low humidity, but most often low or high cloud coverage, respectively. In addition, it's also easier to see that the forecast for Chile is mostly medium humidity and high cloud coverage. A puzzling observation is that Turks and Caicos always has medium humidity, but either low or high cloud coverage.  

Next, I'll look at how location affects minimum and maximum daily temperature using boxplots.  


```{r maxtemp boxplot}
max.means <- weather %>% group_by(location) %>% summarise(average = mean(maxtemp)) 

g <- ggplot(weather, aes(x = location, y = maxtemp))

g + geom_boxplot(fill = "grey") + geom_point(max.means, mapping = aes(x = location, y = average), color = "purple") + geom_line(max.means, mapping = aes(x = location, y = average, group = 1), color = "purple") + labs(title = "8 Day Forecast: Maximum Daily Temperature",x = "Location (highest to lowest latitude)", y = "Maximum Daily Temperature (C)") + coord_flip()
```


```{r mintemp boxplot}
min.means <- weather %>% group_by(location) %>% summarise(average = mean(mintemp))

g1 <- ggplot(weather, aes(x = location, y = mintemp))

g1 + geom_boxplot(fill = "grey") + geom_point(min.means, mapping = aes(x = location, y = average), color = "green") + geom_line(min.means, mapping = aes(x = location, y = average, group = 1), color = "green") + labs(title = "8 Day Forecast: Minimum Daily Temperature", x = "Location (highest to lowest latitude)", y = "Minimum Daily Temperature (C)") + coord_flip()
```

Both boxplots are fairly consistent with each other, showing Turks and Caicos as having the highest minimum and maximum temperature, and locations at the highest and lowest latitudes as having the lowest minimum and maximum temperatures. Although Cucata, Colombia is closer to the equator than Turks and Caicos Islands, a possible reason as to why it may not be warmer is its elevation at 320 meters (1,050 ft).   

Now we'll turn to exploring `humidity` as a quantitative variable, creating a histogram with density plots of `humidity` facetted by `location`.  We'll also create a boxplot of the same information.  


```{r humidity hist and density plot}
g2 <- ggplot(weather, aes(x = humidity))

g2 + geom_histogram(bins = 20, aes(y = ..density..)) + geom_density(color = "aquamarine", fill = "aquamarine") + facet_wrap(~location) + labs(title = "8 Day Forecast: Histogram of Humidity by Location", x = "Humidity (%)")
```



```{r humidity boxplot}
hum.means <- weather %>% group_by(location) %>% summarise(average = mean(humidity))

g3 <- ggplot(weather, aes(x = location, y = humidity))

g3 + geom_boxplot(fill = "grey") + geom_point(hum.means, mapping = aes(x = location, y = average), color = "red") + geom_line(hum.means, mapping = aes(x = location, y = average, group = 1), color = "red") + labs(title = "8 Day Forecast: Average Daily Humidity by Location", x = "Location (highest to lowest latitude)", y = "Average Daily Humidity (C)") + coord_flip()
```

This histogram and boxplot are very helpful in displaying how humidity varies by location. While Argentina and Maine consistently show lower humidity levels, Turks and Caicos is vary consistent in showing high humidity, while Colombia and Chile's humidities are less consistent.


To prepare for the next plot, I'll convert all variables with timestamps to numeric variables of time since 00:00:00, calculated in minutes.  

```{r}
weather$Sunrise <- round(60 * 24 * as.numeric(times(weather$Sunrise)),digits = 2)

weather$Sunset <- round(60 * 24 * as.numeric(times(weather$Sunset)),digits = 2)

weather$Moonrise <- round(60 * 24 * as.numeric(times(weather$Moonrise)), digits = 2)

weather$Moonset <- round(60 * 24 * as.numeric(times(weather$Moonset)), digits = 2)

weather
```


The scatterplot below allows us to look at how sunrise time changes over the course of 8 days in each location.  

```{r sunrise scatterplot}
g4 <- ggplot(weather, aes(x = Day, y = Sunrise))

g4 + geom_point(aes(color = location)) + scale_color_discrete(name = "Location") + labs(title = "8 Day Forecast for Sunrise Time by Location", x = "Day of the Month", y = "Sunrise Time (minutes after 12am)")
```

I think this plot is especially cool to look at, since one can clearly distinguish which locations are in the southern hemisphere, and which are in the northern hemisphere. While the sun is rising increasingly later in the northern hemisphere, it's rising increasingly earlier in the southern hemisphere.  


Lastly, we'll explore the relationship between maximum daily temperature and daily pressure in the scatterplot below. 


```{r maxtemp vs pressure scatterplot}
g5 <- ggplot(weather, aes(x = maxtemp, y = pressure))

g5 + geom_point(aes(color = location)) + geom_smooth(method = lm, formula = y~poly(x,2), color = "black") + scale_color_discrete(name = "Location") + labs(title = "8 Day Forecast: Maximum Daily Temperature vs. Daily Pressure", x = "Maximum Daily Temperature (C)", y = "Daily Pressure (millibars)")
```

There seems to be a medium-strength, positive relationship between temperature and pressure. However, the data from the Turks and Caicos Islands appears to deviate slightly from the general trend, which is why I decided to fit a quadratic model. To fit a more accurate model, we may need to pull data from more locations from the API.  


## Final Thoughts  

Extracting data from an API follows a generally simple process, although different APIs have varying syntax for how to assemble the URL. Hopefully, this vignette was helpful in exemplifying how to access a typical API, clean the data, and perform a few basic analyses.  

To read my blog post about this project, visit <http://atbiggie.github.io>.